---
title: "Udacity - EDA using R"
output:
  html_document:
    fig_height: 7
    fig_width: 9
    dpi: 144
    theme: cerulean
    toc: yes
---
Exercises from Udacity's Exploratory Data Analysis in R MOOC


###Set up packages, chart themes etc

```{r Set up packages and environment, message=FALSE, echo=TRUE, warning=FALSE}

#set up chart theme function and load required packages
chart.theme.size = 14
#theme_set(theme_minimal(chart.theme.size)) #minimal theme as default
source("../r_config/r_env_setup.R") #load packages and cutom ggplot theme function and presets
primary_color = color_primary_pal
highlight_color = color_highlight_pal
set_ggplot_theme01 <- do.call(chart_theme_minimal, as.list(chart_format_default))



```

###Reddit dataset
```{r reddit data, message=FALSE, echo=TRUE, warning=FALSE}

#load the reddit dataset
reddit_df <- read.csv("../data/reddit.csv", header = T,
                      stringsAsFactors=TRUE, sep=",",
                      nrow = 1000)

#Number of rows and columns
dim(reddit_df)

#List of variables
names(reddit_df)

#structure of the dataset
str(reddit_df)

#levels of age range
levels(reddit_df$age.range)

#count summary of age range
table(reddit_df$age.range)

#barblot for age range. Selective coloring of age ranges where count > 300
ggplot(data = reddit_df, aes(x = age.range)) +
  #geom_bar_custom +
  geom_bar(aes(fill = ..count.. > 300)) +
  scale_fill_brewer(palette="Set1")  +
  #geom_hline(yintercept=0, size=0.75, color="grey") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) +
  chart_theme_bar +
  chart_footer 
chart_footnote("Source: Sample Reddit data\nNote*: <add text>")

#p1<-
ggplot(data = reddit_df, aes(x = age.range)) +
  geom_bar_custom +
  #geom_hline(yintercept=0, size=0.75, color="grey") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) +
  chart_theme_bar +
  chart_footer
chart_footnote("Source: Sample Reddit data\nNote*: <add text>")


#ggsave("random_data.png", g, width=12, height=9)


#Notice that under18 bucket is at the end which is non-intuitive
```

###Order a factor variable
```{r create ordered factor variable age.range, message=FALSE, echo=TRUE, warning=FALSE}

reddit_df$age.range <- ordered(reddit_df$age.range, 
                             levels=c("Under 18", "18-24", "25-34", "35-44", "45-54", 
                                      "55-64", "65 or Above" ))

#or alternatively
reddit_df$age.range <- factor(reddit_df$age.range, 
                               levels=c("Under 18", "18-24", "25-34", "35-44", "45-54", 
                                        "55-64", "65 or Above" ),
                               ordered = TRUE)

#barblot for age range after ordering the factor variable
ggplot(data = reddit_df, aes(x = age.range)) +
  do.call(chart_theme_minimal, as.list(chart_format_bar2)) +
  geom_bar(colour = primary_color[1], fill = primary_color[1], alpha = 0.5, size = 0.25) +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) +
  geom_hline(yintercept=0, size=0.75, color="grey") 
#Notice that under18 bucket is placed as the first bar now
```

###Pseudo Facebook Dataset
```{r Load pseudo facebook dataset, message=FALSE, echo=TRUE, warning=FALSE}
#read the dataset into R
fb_df <- read.csv("../data/pseudo_facebook.tsv", header = TRUE,
                  stringsAsFactors = FALSE, sep="\t")

#take a peek into the fb data
glimpse(fb_df)
```

###Create custom color scale for factor variables
```{r Custom color scale, message=FALSE, echo=TRUE, warning=FALSE}

#Create a custom color scale
myColors <- brewer.pal(3,"Set1")

fb_df$gender = factor(fb_df$gender)

#assign factor levels as names to the custom color scale
names(myColors) <- levels(fb_df$gender)

#create an alias for the custom color / fill scale
colScalegender <- scale_colour_manual(name = "gender",values = myColors)
fillScalegender <- scale_fill_manual(name = "gender",values = myColors)
```

##Facebook data - EDA
```{r pseudo facebook dataset, message=FALSE, echo=TRUE, warning=FALSE}


#histogram of b'days
ggplot(data = fb_df, aes(x = dob_day)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1], alpha = 0.4, size = 0.25) +
  xlab("Day of birth") + ylab("# of members") +
  ggtitle("Member Distribution by Day of Birth") +
  scale_x_discrete(breaks = seq(1,31, 2)) +
  scale_y_continuous(labels = comma) +
  chart_footer +
  geom_hline(yintercept=0, size=0.75, color="grey") +
  geom_vline(xintercept=1, size=0.75, color = primary_color[1]) +
  annotate("text", x = 2, y = 6500, 
           label = 
             "Abnormally high counts for 1st day of the month\nSeveral member may be selecting the first\ndropdown choice when selecting the day of birth", 
           size = text.size,color = primary_color[1],
           hjust = 0) 
chart_footnote()

#faceted by month
#histogram of b'days
ggplot(data = fb_df, aes(x = dob_day)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1], alpha = 0.4, size = 0.25) +
  xlab("Day of birth") + ylab("# of members") +
  ggtitle("Member Distribution by Day of Birth faceted by Month") +
  scale_y_continuous(labels = comma) +
  scale_x_discrete(breaks = seq(1,31, 8)) +
  geom_hline(yintercept=0, size=0.75, color="grey")+
  facet_wrap(~dob_month, ncol = 4)
chart_footnote()

#faceted by month and using density (area under each plot is 1) and free scales
#histogram of b'days
ggplot(data = fb_df, aes(x = dob_day)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(aes(y=..density..), colour = "white", 
                 fill = primary_color[1], alpha = 0.4, size = 0.25) +
  scale_y_continuous(labels = percent) +
  xlab("Day of birth") + ylab("# of members") +
  ggtitle("Member Distribution by Day of Birth faceted by Month") +
  scale_x_discrete(breaks = seq(1,31, 8)) +
  geom_hline(yintercept=0, size=0.75, color="grey")+
  facet_wrap(~dob_month, scales = "free", ncol = 4)


#Histogram of friend count (notice that data is concentrated at Zero with few outliers)
ggplot(data = fb_df, aes(x=friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1], alpha = 0.4, size = 0.25) +
  scale_y_continuous(labels = comma) + scale_x_continuous(labels = comma) 

#Friend count after zooming into the previous chart using coord_cartesian
ggplot(data = fb_df, aes(x=friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1], alpha = 0.4, size = 0.25) +
  scale_y_continuous(labels = comma) + 
  scale_x_continuous(labels = comma) +
  coord_cartesian(xlim= c(0, 1000))

#Friend count after limiting the data using limits
ggplot(data = fb_df, aes(x=friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1], alpha = 0.4, size = 0.25) +
  scale_y_continuous(labels = comma) + 
  scale_x_continuous(labels = comma, limits= c(0, 1000)) 


#Friend count after limiting the data using limits and binwidth = 25, and breaks at 50
ggplot(data = fb_df, aes(x=friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(binwidth = 25, colour = "white", fill = primary_color[1], 
                 alpha = 0.4, size = 0.25) +
  scale_y_continuous(labels = comma) + 
  scale_x_continuous(labels = comma, 
                     limits= c(0, 1000), breaks = seq(0, 1000, 100)) 


#Friend count faceted by gender
ggplot(data = fb_df, aes(x=friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(binwidth = 25, colour = "white", fill = primary_color[1], 
                 alpha = 0.4, size = 0.25) +
  scale_y_continuous(labels = comma) + 
  scale_x_continuous(labels = comma, 
                     limits= c(0, 1000), breaks = seq(0, 1000, 300)) +
  facet_wrap(~gender)

#Friend count fill by gender
ggplot(data = filter(fb_df, !is.na(gender)),  # remove records where gender is NA
       aes(x=friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_density(aes(group = gender, fill = gender, colour = gender),
                 binwidth = 25, #colour = "white", #fill = primary_color[1], 
                 alpha = 0.1, size = 0.25) +
  scale_y_continuous() + 
  scale_x_continuous(labels = comma, 
                     limits= c(0, 1000), breaks = seq(0, 1000, 100))


#Friend count fill by gender using specified color palettes
ggplot(data = na.omit(fb_df),  # remove records where any field is NA
       aes(x=friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_density(aes(group = gender, fill = gender, colour = gender),
                 binwidth = 25, #colour = "white", #fill = primary_color[1], 
                 alpha = .2, size = .25) +
  #scale_fill_manual(values=highlight_color) +  # for manually defined palette
  #scale_color_manual(values=highlight_color) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous() + 
  scale_x_continuous(labels = comma, 
                     limits= c(0, 1000), breaks = seq(0, 1000, 100))


```


```{r summary of friend count by gender,message=FALSE, echo=TRUE, warning=FALSE}


#count of levels in gender
fb_df %>%
  group_by(gender) %>%
  summarise(count = n())

#min value of each variable by gender 
fb_df %>%
  na.omit() %>% #filter out rows with missing values in any variable
  group_by(gender) %>%
  summarise_each(funs(min))


```


```{r Histogram for Tenure,message=FALSE, echo=TRUE, warning=FALSE}

#Histogram for tenure
ggplot(data = fb_df, aes(x = tenure/365)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(binwidth = 1/4, colour = "white", fill = primary_color[1], 
                 alpha = .3, size = .25) +
  xlab("Tenure (years)") +
  ylab("# of Facebook members") +
  ggtitle("Distribution of members by tenure") +
  scale_y_continuous(labels = comma) + 
  scale_x_continuous(labels = comma, 
                     limits= c(0, 7), breaks = seq(0, 7, 1)) +
  geom_hline(yintercept=0, size=0.75, color="grey") +
  annotate("text", x = Inf, y = -Inf, label = "sumitbajaj.me",
           hjust=1.1, vjust= -.5, col="gray", cex=4, alpha = 0.8)

```

```{r Histogram for Age, message=FALSE, echo=TRUE, warning=FALSE}

#histogram for user age
ggplot(data = fb_df, aes(x = age)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1],
                 alpha = 0.3, size = .25) +
  geom_hline(yintercept=0, size = 0.75, color = "grey") +
  scale_y_continuous(labels = comma) +
  ggtitle("Distribution of members by Age") +
  xlab("member age") +
  ylab("# of Facebook members")


#histogram for user age with adjusted binwidth
#A binwidth of 1 allows us to visualize any unusual spikes in the data
ggplot(data = fb_df, aes(x = age)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1],
                 alpha = 0.3, size = .25,
                 binwidth = 1) +
  geom_hline(yintercept=0, size = 0.75, color = "grey") +
  scale_y_continuous(labels = comma) +
  ggtitle("Distribution of members by Age") +
  xlab("member age") +
  ylab("# of Facebook members") +
  scale_x_discrete(labels = comma, breaks = seq(min(fb_df$age), max(fb_df$age), 10))

```

### Transforming data - Log, Sqrt etc
```{r Transforming data, message=FALSE, echo=TRUE, warning=FALSE }

summary(fb_df$friend_count)

#Log transform for the # of likes
summary(log10(fb_df$friend_count))

#Log transform for the # of likes +1 to avoid infinity at log zero
summary(log10(fb_df$friend_count + 1))

# visualize friend count 
p1 <-
ggplot(data = fb_df, aes(x = friend_count)) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1],
                 alpha = 0.3, size = .25) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma)

#visualize friend count with log
p2 <-
ggplot(data = fb_df, aes(x = log10(friend_count + 1))) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1],
                 alpha = 0.3, size = .25) +
  scale_y_continuous(labels = comma)

#visualize friend count with square root
p3 <-
ggplot(data = fb_df, aes(x = sqrt(friend_count))) +
  do.call(chart_theme_minimal, as.list(chart_format_hist_no_vgrid)) +
  geom_histogram(colour = "white", fill = primary_color[1],
                 alpha = 0.3, size = .25) +
  scale_y_continuous(labels = comma)

grid.arrange(p1, p2, p3, ncol = 1)

#alternate way for log scale P2. Note that the x axis label doesn't reflect log
p1 + scale_x_log10()

```

###Frequency polygons for friend_count
Note that sum(..count..) will sum across color, so the percentages displayed are percentages of total users. To plot percentages within each group, you can try y = ..density...

```{r Frequency polygons with manular color and fill scales, message=FALSE, echo=TRUE, warning=FALSE}


ggplot(data = filter(fb_df, !is.na(gender)), 
       aes(x = friend_count)) +
  do.call(chart_theme_minimal, as.list((chart_format_hist_no_vgrid["fsize"] = 14))) + 
  geom_histogram(aes(fill = gender),
                 binwidth = 10,
                 color = 'white',
                 alpha = 0.5, size = 0.25) +
  fillScalegender +
  scale_x_continuous(labels = comma,
                     lim = c(0, 1000)) +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "right")

# frequency polygon - friend count
ggplot(data = filter(fb_df, !is.na(gender)), 
       aes(x = friend_count,
           y = ..count../sum(..count..))) +
  do.call(chart_theme_minimal, as.list((chart_format_hist_no_vgrid["fsize"] = 14))) +
  geom_freqpoly(aes(color = gender, fill = gender),
                binwidth = 10,
                alpha = 1, size = 1.5) +
  colScalegender +
  #scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(labels = comma,
                     lim = c(0, 1000)) +
  theme(legend.position = "right")


```

###Feature usage - Likes males vs females
```{r likes male vs female, message=FALSE, echo=TRUE, warning=FALSE}

# frequency polygon - www_likes
p1 <-
ggplot(data = filter(fb_df, !is.na(gender)), 
       aes(x = www_likes)) +
  do.call(chart_theme_minimal, as.list((chart_format_hist_no_vgrid["fsize"] = 14))) +
  geom_freqpoly(aes(color = gender, fill = gender),
                #binwidth = 10,
                alpha = 1, size = .5) +
  colScalegender +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "right")

print(p1)

p1 + scale_x_log10() +
  ylab("Log10Count")

#Who gets more likes males or females?
by(fb_df$www_likes, fb_df$gender, sum)

fb_df %>%
  group_by(gender) %>%
  summarise(num_www_likes = sum(www_likes),
            n=n(),
            likes_per_person = num_www_likes/n()) 

#Ans: Females get more likes overall and average female get more than 3x the # of male likes

#table <- xtable(freq,floating=FALSE)
#print(table, type = "HTML")


```

###Boxplots - friend_count
```{r Boxplots Friend Count, message=FALSE, echo=TRUE, warning=FALSE}

b1 <-
ggplot(data = filter(fb_df, !is.na(gender)), 
       aes(x = gender, y = friend_count)) +
  geom_boxplot() +
  do.call(chart_theme_minimal, as.list((chart_format_hist_no_vgrid["fsize"] = 14))) +
  scale_y_continuous(labels = comma) +
  xlab("")

print(b1)

#zoom to focus on users having 0-1000 friends
b1 +
  coord_cartesian(ylim = c(0, 1000))

#zoom further focus on users having 0-250 friends
b1 +
  coord_cartesian(ylim = c(0, 250))

#summary of friend_count by gender
by(fb_df$friend_count, fb_df$gender, summary)

```

###Logical operators for flag variables
At times variables may have very sparse data e.g. # of mobile logins. In such cases, you may want to create a variable that just captures whether a given feature was ever used.

```{r Logical operators, message=FALSE, echo=TRUE, warning=FALSE}

#Summary of mobile likes
summary(fb_df$mobile_likes)

#You can observe that more than 25% of users have zero mobile likes
summary(fb_df$mobile_likes > 0)
#More than 35% of users have no mobile likes

#mobile checkin flag variable creation
fb_df <- fb_df %>%
  mutate(mobile_checkin = NA) %>%
  mutate(mobile_checkin = ifelse(mobile_likes >0, 1, 0)) %>%
  mutate(mobile_checkin = factor(mobile_checkin))

summary(fb_df$mobile_checkin)

#What % of people ever checked in using mobile
print(sum(fb_df$mobile_checkin == 1) / length(fb_df$mobile_checkin))

```

##Diamonds dataset EDA
###Diamonds- Basic dataset info
```{r Diamonds dataframe, message=FALSE, echo=TRUE, warning=FALSE}
dim(diamonds)
glimpse(diamonds)
```
###Diamonds- EDA
```{r Diamonds EDA, message=FALSE, echo=TRUE, warning=FALSE}
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(colour = "white", fill = primary_color[1], alpha = 0.4, size = 0.25) +
  geom_vline(xintercept=mean(diamonds$price), size=0.75, color = "grey") +
  annotate("text", x = mean(diamonds$price)*1.03, y = 15000, 
           label = "mean", hjust = 0) +
  geom_vline(xintercept=median(diamonds$price), size=0.75, color = "grey") +
  annotate("text", x = median(diamonds$price)*1.03, y = 12000, 
           label = "median", hjust = 0)

summary(diamonds$price)

#count by price points
diamonds %>%
  mutate(price_pnt = ifelse(price < 250, "<250", 
                            ifelse(price < 500, "<500",
                                   ifelse(price >= 15000, ">15k", "other")))) %>%
  group_by(price_pnt) %>%
  summarise(n = n())

```

##Facebook - Explore two variables
```{r scatter plots, message=FALSE, echo=TRUE, warning=FALSE}

#scatterplot - relation between age and friend count
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  set_ggplot_theme01 +
  geom_point(alpha = 1,
             color = primary_color[1])

#set alpha to 1/20 i.e 20 data points make one solid dot
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  set_ggplot_theme01 +
  geom_point(alpha = 1/20,
             color = primary_color[1])

#limit the age from 13 to 90 yrs
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  set_ggplot_theme01 +
  geom_point(alpha = 1/20,
             color = primary_color[1]) +
  xlim(13, 90)

#add jitter to avoid straight vertical lines
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  set_ggplot_theme01 +
  geom_jitter(alpha = 1/20,
              color = primary_color[1]) +
  scale_x_continuous(limits = c(13, 90), breaks = seq(13, 90, 10))

#add square root Y axis to reduce right skew
#limit the age from 13 to 90 yrs
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  set_ggplot_theme01 +
  geom_jitter(alpha = 1/20,
              position = position_jitter(h = 0),
              color = primary_color[1]) +
  scale_x_continuous(limits = c(13, 90), breaks = seq(13, 90, 10)) +
  coord_trans(y = "sqrt")

```
The square root, x to x^(1/2) = sqrt(x), is a transformation with a
moderate effect on distribution shape: it is weaker than the logarithm
and the cube root. It is also used for reducing right skewness, and also
has the advantage that it can be applied to zero values. Note that the
square root of an area has the units of a length. It is commonly applied
to counted data, especially if the values are mostly rather small.

If we add noise to zero we could end up with -ve numbers where sqrt will be imaginary. Set the position parameter equal to position_jitter and pass it a min height of zero.

###Summary metrics on charts
```{r summary metrics on charts, message=FALSE, warning=FALSE, echo=TRUE}

ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  set_ggplot_theme01 +
  geom_jitter(alpha = 1/20,
              position = position_jitter(h = 0),
              color = "light grey") +
  scale_x_continuous(limits = c(13, 90), breaks = seq(13, 90, 10)) +
  coord_trans(y = "sqrt") +
  geom_line(aes(color = "Mean"), stat = "summary", fun.y = mean) +
  geom_line(aes(color = "10% Quantile"), stat = "summary", fun.y = quantile, 
            probs = 0.1, linetype = 2) +
  geom_line(aes(color = "90% Quantile"), stat = "summary", fun.y = quantile, 
            probs = 0.9, linetype = 2) +
  #scale_color_manual(values=primary_color[1]) +
  #annotate("text", x=max(fb_df$age)+1, y = mean(fb_df$friend_count), label = "mean", hjust = 0) +
  theme(legend.position = "right") +
  guides(colour = guide_legend(override.aes = list(size=1)))


#zoom in using coord-cartesian 
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  set_ggplot_theme01 +
  coord_cartesian(xlim = c(13, 70), ylim = c(0, 1000)) +
  geom_jitter(alpha = 1/20,
              position = position_jitter(h = 0),
              color = "light grey") +
  #scale_x_continuous(limits = c(13, 90), breaks = seq(13, 90, 10)) +
  #coord_trans(y = "sqrt") +
  geom_line(aes(color = "Mean"), stat = "summary", fun.y = mean) +
  geom_line(aes(color = "10% Quantile"), stat = "summary", fun.y = quantile, 
            probs = 0.1, linetype = 2) +
  geom_line(aes(color = "90% Quantile"), stat = "summary", fun.y = quantile, 
            probs = 0.9, linetype = 2) +
  #scale_color_manual(values=primary_color[1]) +
  #annotate("text", x=max(fb_df$age)+1, y = mean(fb_df$friend_count), label = "mean", hjust = 0) +
  theme(legend.position = "right") +
  guides(colour = guide_legend(override.aes = list(size=1)))

```
###Correlation
A correlation of >0.3 is mild, >0.5 is moderate and >0.7 is pretty strong

```{r Correlation, message=FALSE, warning=FALSE, echo=TRUE}

cor.test(fb_df$age, fb_df$friend_count, method = "pearson")

#alternatively for less verbose code
with(fb_df, cor.test(age, friend_count, method = "pearson"))

with(filter(fb_df, age<=70), cor.test(age, friend_count, method = "pearson"))

#spearman coeff for monotonic relationships
with(filter(fb_df, age<=70), cor.test(age, friend_count, method = "spearman"))

```

###Strong correlations

engagement variables all measure essentiall the same thing - engagement. e.g. # of photo uploads, no. of posts etc..
```{r Strong Correlations, message=FALSE, warning=FALSE, echo=TRUE}

ggplot(data = fb_df, aes(x = www_likes_received, y = likes_received)) +
  set_ggplot_theme01 +
  geom_point(position = position_jitter(h=0),
             color = "light grey",
             alpha = 1/20) +
  xlim(0, quantile(fb_df$www_likes_received, 0.95)) +
  ylim(0, quantile(fb_df$likes_received, 0.95)) +
  geom_smooth(method = "lm", color = primary_color[1], size = 2) +
  annotate("text", x = Inf, y = -Inf, label = "sumitbajaj.me",
           hjust=1.1, vjust=-.5, col="gray", cex=4, alpha = 0.8)

#correlation coeff
cor.test(fb_df$www_likes_received, fb_df$likes_received, method = "pearson")

```

###More caution with correlation
```{r Caution with correlation, message=FALSE, warning=FALSE, echo=TRUE}
#load the mitchel dataset for soil temp study
#install.packages("alr3")
library(alr3)
glimpse(Mitchell)

#Range of the month variable
range(Mitchell$Month)

ggplot(data = Mitchell, aes(x = Month, y = Temp)) +
  chart_theme01 +
  geom_point() +
  chart_footer+
  scale_x_continuous(breaks = seq(0, 203, 12))  #increments of 12 months


with(Mitchell, cor.test(Month, Temp, method = "pearson"))

#non parametric test
with(Mitchell, dcor.ttest(Month, Temp))


#notice the cyclical patterns with line chart and aspect ratio using coord_fixed (y/x)
ggplot(data = Mitchell, aes(x = Month, y = Temp)) +
  chart_theme01 +
  geom_point() +
  geom_line(color = color_primary_pal[1], alpha = 0.3, size = 2) +
  coord_fixed(ratio = 1) +
  chart_footer +
  scale_x_continuous(breaks = seq(0, 203, 12))  #increments of 12 months


#to compare monthly trends
ggplot(data = Mitchell, aes(x = Month%%12, y = Temp)) +
  chart_theme01 + 
  geom_point(color = color_primary_pal[1], alpha = 0.5, size = 4) +
  scale_x_continuous(breaks = seq(0, 13, 1)) +
  chart_footer 

```
###Noise and smootheming
```{r Noise and smoothening, message=FALSE, warning=FALSE, echo=TRUE}

pf.fc_by_age <- fb_df %>%
  filter(age < 71) %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n())

#plot mean friend count against age (yrs)
p1 <-
ggplot(data = pf.fc_by_age, aes(x = age, y = friend_count_mean)) +
  chart_theme01 +
  scale_y_continuous(limit = c(0,400)) +
  scale_x_continuous(limit = c(0, 72)) +
  geom_line(color = color_primary_pal[1], alpha = 1, size = 1) +
  geom_smooth(size = 0.1, fill = grey(0.9)) +
  chart_footer

pf.fc_by_age[15:20,]


#mean friend count by age (year.month) to generate an even noisier plot
pf.fc_by_age_months <- fb_df %>%
  filter(age < 71) %>%
  mutate(age_with_months = age + (12 - dob_month)/12) %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n())


#plot mean friend count against age.month
p2 <-
ggplot(data = pf.fc_by_age_months, aes(x = age_with_months, y = friend_count_mean)) +
  chart_theme01 +
  scale_y_continuous(limit = c(0,400)) +
  scale_x_continuous(limit = c(0, 72)) +
  geom_line(color = color_primary_pal[2], alpha = 1, size = 0.25) +
  geom_smooth(size = 0.1, fill = grey(0.9)) +
  chart_footer

#Loess regression to smooth the mean friend count
p3 <- 
  fb_df %>%
  filter(age < 71) %>%
  ggplot(aes(x = round(age/5)*5, y = friend_count)) +
  chart_theme01 +
  geom_line(stat = "summary", fun.y = mean,
            color = color_primary_pal[3], alpha = 1, size = 1) +
  scale_x_continuous(limit = c(0, 72)) +
  coord_cartesian(ylim = c(0, 400)) +
  geom_smooth(size = 0.1, fill = grey(0.9)) +
  chart_footer
  
grid.arrange(p2, p1, p3, ncol = 1)

```

##More than two variables for EDA
###Third qualitative variable
```{r Third qualitative variable, message=FALSE, warning=FALSE, echo=TRUE}

ggplot(data = subset(fb_df, !is.na(gender)), aes(x = age, y = friend_count)) +
  chart_theme01 +
  geom_line(stat = "summary", fun.y = mean,
            aes(color = gender)) +
  #colScalegender +
  scale_colour_brewer(palette="Set1") +
  legend_top +
  legend_title_hide +
  legend_size_override() +
  chart_footer +
  ggtitle("Mean friend count by age and gender")

#mean friend count by age>genger
pf.fc_by_age_gender <-
  fb_df %>%
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count = mean(friend_count),
                          median_friend_count = median(as.numeric(friend_count)),
                          n = n())

head(pf.fc_by_age_gender, 4)

#median friend count by age for each gender
ggplot(data = pf.fc_by_age_gender, aes(x = age, y = median_friend_count)) +
  chart_theme01 +
  geom_line(aes(color = gender)) +
  #colScalegender +
  scale_colour_brewer(palette="Set1") +
  legend_top +
  legend_title_hide +
  legend_size_override() +
  chart_footer +
  ggtitle("Median friend count by age and gender")
```

##Reshape from long to wide format
```{r Reshape data using Tidyr, message=FALSE, warning=FALSE, echo=TRUE}

#spread to wide format key-value
pf.fc_by_age_gender_wide <- pf.fc_by_age_gender %>%
  select(1, 2, 4) %>%
  spread(gender, median_friend_count) %>%
  mutate(ratio = male/female)

head(pf.fc_by_age_gender_wide, 3)

#plot ratio of median friends female / male
ggplot(data = pf.fc_by_age_gender_wide, aes(x = age, y = female/male)) +
  chart_theme01 +
  geom_line(color = color_primary_pal[2], alpha = 1, size = 0.75) +
  coord_cartesian(ylim = c(0, 3)) +
  geom_hline(yintercept =1, linetype = 2, size = 1, color = color_highlight_pal[1]) +
  chart_footer
#note: one possible reason. More growth from newer countries. Initial users more likely to be male.

```

##Add bucketed variables like year joined using cut
```{r Add more variables like year joined, message=FALSE, warning=FALSE, echo=TRUE}
#create variable for year joined fb
fb_df$year_joined = 2014 - ceiling(fb_df$tenure/365)

summary(fb_df$year_joined)
#looks like most people joined in last 2 years

table(fb_df$year_joined)
        
#create bucketed variable for year joined
fb_df <- fb_df %>%
  mutate(year_joined_bucket = cut(year_joined, c(2004, 2009, 2011, 2012, 2014)))

table(fb_df$year_joined_bucket)

#plot each year_join bucket for median friend count against age
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  chart_theme01 +
  chart_footer +
  geom_line(aes(color = year_joined_bucket),
            stat = "summary", fun.y = median) +
  scale_colour_brewer(palette="Set1") +
  legend_top +
  legend_size_override()


#Plot the grand mean as y intercept
#plot each year_join bucket for median friend count against age
ggplot(data = fb_df, aes(x = age, y = friend_count)) +
  chart_theme01 +
  chart_footer +
  geom_line(aes(color = year_joined_bucket),
            stat = "summary", fun.y = mean) +
  scale_colour_brewer(palette="Set1") +
  geom_line(stat = "summary", fun.y = mean, linetype = 2, aes(color = "Grand Mean")) +
  annotate("text", x = 15, y = mean(fb_df$friend_count)*1.2, label = "Grand Mean") +
  legend_top +
  legend_size_override()


#Friendships initiated by tenure
ggplot(data = filter(fb_df, !is.na(tenure), !is.na(friendships_initiated), tenure >1, friendships_initiated >1), 
       aes(x = tenure, y = friendships_initiated/tenure)) +
  geom_line(aes(color = year_joined_bucket),
            stat = "summary", fun.y = mean) +
  chart_theme01 +
  chart_footer +
  legend_top + legend_size_override() 

#It appears that users with higher tenure initiate less friendships  

#Smoothening the tenure - bias variance tradeoff. as we increase the binwidth, the variance goes down and the bias goes up
ggplot(data = filter(fb_df, !is.na(tenure), !is.na(friendships_initiated), tenure >1, friendships_initiated >1), 
       aes(x = round(tenure/7)*7, y = friendships_initiated/tenure)) +
  geom_line(aes(color = year_joined_bucket),
            stat = "summary", fun.y = mean) +
  chart_theme01

ggplot(data = filter(fb_df, !is.na(tenure), !is.na(friendships_initiated), tenure >1, friendships_initiated >1), 
       aes(x = round(tenure/30)*30, y = friendships_initiated/tenure)) +
  geom_line(aes(color = year_joined_bucket),
            stat = "summary", fun.y = mean) +
  chart_theme01

ggplot(data = filter(fb_df, !is.na(tenure), !is.na(friendships_initiated), tenure >1, friendships_initiated >1), 
       aes(x = round(tenure/90)*90, y = friendships_initiated/tenure)) +
  geom_line(aes(color = year_joined_bucket),
            stat = "summary", fun.y = mean, , size = 1) +
  chart_theme01


#using geom_smooth 
ggplot(data = filter(fb_df, tenure >=1), 
       aes(x = tenure, y = friendships_initiated/tenure)) +
  geom_smooth(aes(color = year_joined_bucket)) +
  chart_theme01 +
  scale_x_continuous(labels = comma)
```

##Yogurt purchase data EDA
```{r Yogurt purchase data EDA, message=FALSE, warning=FALSE, echo=TRUE}
yo <- read.csv("../data/yogurt.csv", header = T, stringsAsFactors = TRUE, sep = ",")
head(yo)
str(yo)

#change id to factor variable from int
yo$id <- as.factor(yo$id)

#Price histogram
ggplot(data = yo, aes(x = price)) +
  chart_theme01 +
  geom_histogram(binwidth = 1, fill = color_primary_pal[1])
#the price histogram indicates discreteness

length(unique(yo$price))
#only 20 unique prices

#new variable for total # of purchases in a transaction
yo <- yo %>%
  mutate(all.purchases = strawberry + blueberry	+ pina.colada	+ plain	+ mixed.berry)
summary(yo$all.purchases)

#histogram of all.purchases
ggplot(data = yo, aes(x = all.purchases))+
  chart_theme01 +
  geom_histogram(fill = color_primary_pal[1], binwidth = 1)

#scatterplot of price vs time
ggplot(data = yo, aes(x = time, y = price))+
  chart_theme01 +
  geom_point(color = color_primary_pal[1], alpha = 0.5)

#the most common prices seem to be increasing over time
#the scattered lower prices could be due to markdowns or usage of coupons


```

#looking at a sample of yogurt households to understand deeper
```{r sampling yogurt data, message=FALSE, warning=FALSE, echo=TRUE}
#set the seed for reproducible samples
set.seed(4230)

#draw 16 households
sample.ids <- sample(levels(yo$id), 16)

#scatterplot of price vs time
ggplot(data = filter(yo, id %in% sample.ids), aes(x = time, y = price))+
  chart_theme01 +
  geom_point(color = color_primary_pal[1], alpha = 0.5, aes(size = all.purchases)) +
  geom_line(color = color_highlight_pal[1], size = .5, alpha = 0.8) +
  legend_top +
  facet_wrap(~id) +
  ggtitle("# purchases and price trend over time for 16 random households")
  

```

##Scatterplot matrices
```{r scatterplot matrices, message=FALSE, warning=FALSE, echo=TRUE}
#set the seed for reproducible samples
set.seed(1836)

#extract columns 2:5 from fb dataframe
pf_subset <- fb_df[, c(2:5)]

ggpairs(pf_subset[sample.int(nrow(pf_subset), 100), ],
         params=list(corSize=4, fontsize = 4))

```
```{r code snippet, message=FALSE, warning=FALSE, echo=TRUE}

```