---
title: "Dplyr usage examples"
output:
  html_document:
    fig_height: 7
    fig_width: 9
    dpi: 144
    theme: cerulean
    toc: yes
    
---

#Useful links
* http://cran.r-project.org/web/packages/dplyr/vignettes/
* http://bit.ly/wrangling-webinar
* http://rpubs.com/justmarkham/dplyr-tutorial
* http://www.princeton.edu/~mattg/statar/
* http://rpackages.ianhowson.com/cran/dplyr/
* http://stat545-ubc.github.io/bit001_dplyr-cheatsheet.html
* http://stackoverflow.com/questions/28222876/passing-column-name-as-parameter-to-function-in-r-language-dplyr

```{r global_options, include=FALSE}

#http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html
knitr::opts_chunk$set(fig.width=9, 
                      fig.height=7, 
                      fig.path='Figs/',
                      echo=TRUE, 
                      warning=FALSE, 
                      message=FALSE,
                      include = TRUE, #show figures, FALSE will evaluate code but suppress results
                      #fig.show="hide", #if you want to hide figures
                      results = "show"
                      )

knitr::opts_chunk$set(cache=TRUE)

```



#Setup
Set up packages, chart themes, slack notifications etc
```{r Set up packages and environment, cache=FALSE}

#load packages
source("../../r_config/r_packages_setupv1.R") 

#load custom functions
source("../../r_config/r_functions_repository.R")

#load custom ggplot2 theme functions
source("../../r_config/r_visualization_setupv1.R")

#set default theme for charts
theme_set(chart_theme_custom_base(fsize = 18, background_grey = 0, vgrid = 1, hgrid = 1)) 

#data table global options
options(DT.options = list(pageLength = 5, language = list(search = 'Search:')))
#datatable(head(iris))

#setup slack bot for notifications
setup_slack(token_path = "../../../api_tokens/api_tokens - list.csv", channel_name = "#code_status")

#post a test message
slackr("Everything setup!")

```

#dplyr functionality
- Five basic verbs: filter, select, arrange, mutate, summarise (plus group_by)
- Can work with data stored in databases and data tables
- Joins: inner join, left join, semi-join, anti-join
- Window functions for calculating ranking, offsets, and more


#Load flights dataset
```{r load flights data, cache=FALSE, echo=TRUE, message=FALSE, warning = FALSE, eval=TRUE}


#load the NY flights data
library(nycflights13)
nyflights = tbl_df(flights)
nyflights

#load the Houston flights data
library("hflights")

#use tbl_df(local data frame is a wrapper for a df that prints within visible area) to format the df
flights <- tbl_df(hflights)

#view the number of rows and cols
dim(flights)

#glimpse the structure
glimpse(flights)

#view the top few rows
datatable(head(hflights))

#will only print columns that can be fitted on the screen
head(flights, n = 3)

```

#Filter - Keep rows matching the specified criteria
* Command structure for all dplyr verbs
* first argument is a dataframe
* return value is also a dataframe
* nothing is modified in place

```{r dplyr-filter}

#get all flights on January 1

#base R approach: flights[flights$Month == 1 & flights$DayofMonth == 1]
#dplyr approach below
#flights %>% filter(Month ==1, DayofMonth ==1)
#you can use either comma or & for AND condition
flights %>% filter(Month ==1 & DayofMonth ==1)
#we can observe that 552 flights met the specified criteria

#filter flights from AA or UA 
flights %>% filter(UniqueCarrier == "UA" | UniqueCarrier == "AA")
#we can observe that 5,316 flights were from UA or AA

#same result using the %in% operator
flights %>% filter(UniqueCarrier %in% c("AA", "UA"))
```

#Filter - between, slice, sample_n, top_n, distinct

```{r dplyr-filter-additional options}
# `filter` supports multiple conditions
nyflights %>% filter(dep_time >= 600 & dep_time <= 605)

# `between` is a concise alternative for determining if numeric values fall in a range
nyflights %>% filter(between(dep_time, 600, 605))

# `is.na()` can also be used while filtering
nyflights %>% filter(!is.na(dep_time))

# `slice` filters rows by position
nyflights %>% slice(1000:1005)

# keep the first 3 rows within each group
nyflights %>%
  group_by(month, day) %>%
  slice(1:3)

# sample 3 rows from each group
nyflights %>% group_by(month, day) %>% sample_n(size = 3)

# keep 3 rows from each group with the top dep_delay
nyflights %>% group_by(month, day) %>% top_n(n = 3, dep_delay)

# also sort by dep_delay within each group
nyflights %>% group_by(month, day) %>% top_n(n=3, dep_delay) %>% arrange(desc(dep_delay))

# unique rows can be identified using `unique()` from base R
nyflights %>% select(origin, dest) %>% arrange(origin, dest) %>% unique()

# dplyr provides a more efficient alternative `distinct`
nyflights %>% select(origin, dest) %>% distinct()

```


#Select - Choose columns by name
* Similar syntax to `filter`
* Just like SELECT in SQL

```{r dplyr-select}
# base R approach to select DepTime, ArrTime, and FlightNum columns
flights[, c("DepTime", "ArrTime", "FlightNum")]

#dplyr approach
flights %>% select(DepTime, ArrTime, FlightNum)

#use colon `:` to select multiple contiguous column, and use `contains` to match columns by name
flights %>% select(Year:DayofMonth, contains("Taxi"), contains("Delay"))

```

#Select - Rename and other options
```{r dplyr-select-rename}

#list of columns
names(nyflights)

#you can use the `-` sign to hide columns
nyflights %>% select(-month, -day)

#hide a range of columns
nyflights %>% select( -(dep_time : arr_delay) )

#choose columns using a char vector of col names
cols <- c("carrier", "flight", "tailnum")

# `one_of` selects columns specified in a char vector
nyflights %>% select(one_of(cols))

# `select` can also be used to rename cols, though all cols not mentioned are dropped
nyflights %>% select(tail = tailnum)

# `rename` does the same thing, except all cols not mentioned are kept
nyflights %>% rename(tail = tailnum)

```


#Pipes
* Similar to UNIX
* Write commands in the natural order by using the `%>%` operator
* Pipes improve readability when there are many operations
* Can be used to replace nesting in R commands outside of `dplyr` too

```{r dplyr-pipes}

#all flights (carrier, delay) where delay is greather than 60
flights %>% select(UniqueCarrier, DepDelay) %>% filter(DepDelay > 60)


#illustration of pipes - two vectors and computation of Euclidian distance between them
x1 <- 1:5 
x2 <- 2:6
(x1-x2)^2 %>% sum() %>% sqrt()

```

#Arrange - Sort rows
```{r dplyr-arrange}
#base R approach to select UniqueCarrier and DepDelay columns and sort by DepDelay
#prefix the sorting variable by a minus `-` sign to indicate DESCENDING order
flights[order(-flights$DepDelay), c("UniqueCarrier", "DepDelay")]

#dplyr approach to sorting. Use `desc` for descending
flights %>% select(UniqueCarrier, DepDelay) %>% arrange(desc(UniqueCarrier), DepDelay)

```

#Mutate - Add new columns/ variables
* Create new columns that are functions of other existing columns

```{r dplyr-mutate}
#base R approach to create a new variable Speed (in mph)
flights$Speed <- flights$Distance / flights$AirTime*60
flights[ , c("Distance", "AirTime", "Speed")] #all rows with the selected columns in the list

#dplyr `mutate` creates a new variable and keeps all existing variables
flights %>% select(Distance, AirTime) %>% mutate(Speed = Distance/AirTime*60)

# `transmute` only keeps the new variables
nyflights %>% transmute(speed = distance/air_time*60)

# example dataset with row names
mtcars %>% head()

# `add_rownames()` turns row names into an explicit variable
mtcars %>% add_rownames("model") %>% head()

# dplyr no longer prints row names for local data frames
mtcars %>% tbl_df() %>% head()

```

#Summarise - Group values by variable(s)
* Useful with data that has been `grouped by` one or more variables
* `group_by` creates the groups that will be operated upon
* `summarise` uses the specified aggregation function to summarise each group
* `summarise_each` allows you to apply the same summary function to multiple columns at once
* `mutate_each` is also available
* `tally`
* `group_size`
* `n_groups`
* `ungroup`
* Helper function `n()` counts the number of rows in a group
* Helper function `n_distinct(vector)` counts the number of unique items in that vector
* Grouping can sometimes be useful without summarising

```{r dplyr-group_by-summarise}
# base R approaches to calculate the average arrival delay to each destination
head(with(flights, tapply(ArrDelay, Dest, mean, na.rm=TRUE))) #summarise ArrDelay by Dest using "mean"
#or
head(aggregate(ArrDelay ~ Dest, flights, mean)) #summarise ArrDelay by Dest using "mean"

#dplyr approach
flights %>%
  group_by(Dest) %>%
  summarise(avg_delay = mean(ArrDelay, na.rm = TRUE)) %>%
  arrange(desc(avg_delay))


#for each carrier, calculate the % of flights cancelled or diverted
flights %>%
  group_by(UniqueCarrier) %>%
  summarise_each(funs(mean(., na.rm = TRUE)),   #the functions to summarise with
                 Cancelled, Diverted)          #the columns to be reduced to summarized values


#for each carrier, calculate the min and max arrival and departure delays
#when >1 functions and columns are specified, the function name is suffixed to the col name being summarised
flights %>%
  group_by(UniqueCarrier) %>%
  summarise_each(funs(MIN = min(., na.rm=TRUE),  #you can specify the prefix for the summarised columns
                      max(., na.rm=TRUE)),
                 matches("Delay"))  #all columnns containing "Delay"

#for each day of the year, count the total # of flights and sort in descending order
flights %>%
  group_by(Month, DayofMonth) %>%
  summarise(flight_count = n()) %>%
  arrange(desc(flight_count))

#can be simplified by using 'tally' function. Its a wrapper for summarise n()
flights %>%
  group_by(Month, DayofMonth) %>%
  tally(sort = TRUE) #tally sorts in descending order
  
#for each destination, count the total # of flights and the # of distinct planes that flew there
flights %>%
  group_by(Dest) %>%
  summarise(flight_count = n(),   #total num of flights
            plane_count = n_distinct(TailNum)) #unique planes

#tabulate
#for each destination, show the # of cancelled and not cancelled flights
flights %>%
  group_by(Dest) %>%
  select(Cancelled) %>%
  table() %>%
  head()

# tally to count number of flights in each month. alternative to summarise(n = n())
nyflights %>% group_by(month) %>% tally(sort = TRUE)

# or using count
nyflights %>% count(month, sort = TRUE)

# you can sum over a specific variable
nyflights %>% group_by(month) %>% summarise(dist = sum(distance)) %>% head()

# `tally` and `count` have a `wt` parameter for this purpose
nyflights %>% group_by(month) %>% tally(wt = distance) %>% head()
nyflights %>% count(month, wt = distance) %>% head()

# `group_size()` returns the counts as a vector
nyflights %>% group_by(month) %>% group_size() 

# `n_groups()` simply reports the number of groups
nyflights %>% group_by(month) %>% n_groups()

# group by 2 variables, sumarise, arrange (output is possibly confusing)
nyflights %>% group_by(month, day) %>% summarise(cnt = n()) %>% arrange(desc(cnt)) %>% print(n = 35)

# `ungroup()` before arranging to arrange across all groups
nyflights %>% group_by(month, day) %>% summarise(cnt = n()) %>% ungroup() %>% arrange(desc(cnt)) %>% head()

```

#Window Functions
* http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html
* https://danieljhocking.wordpress.com/2014/12/03/lags-and-moving-means-in-dplyr/
* Aggregation functions e.g `mean` take n inputs and return 1 value
* `Window functions` take n inputs and return n values
  * ranking and ordering functions - `min_rank`
  * offset functions - `lead`, `lag`
  * cumulative aggregates - `cummean`

Window functions operate on a partition or "window" of a result set, and return a value for every row in that window. In contrast, nonwindowed functions perform their calculations with respect to every row in the result set. 

Unlike group functions that aggregate result rows, all rows in the table expression are retained.

The values returned are calculated by utilizing values from the sets of rows in that window. The window defines, for each row in the table, a set of rows that is used to compute additional attributes. A window is defined using a window specification (the `OVER` clause) in `SQL`, and is based on 3 main concepts
* Window partitioning, which forms groups of rows (PARTITION clause)
* Window ordering, which defines an order or sequence of rows within each partition (ORDER BY clause)
* Window frames, which are defined relative to each row to further restrict the set of rows (ROWS specification)

```{r dplyr-window functions}
#for each carrier, calculate which 2 days of the year they had their longest departure delays
#note- smallest (not largest) value is ranked as 1, so rem to use `desc` to rank by largest value for delays
flights %>%
  group_by(UniqueCarrier) %>%
  select(Month, DayofMonth, DepDelay) %>%
  filter(min_rank(desc(DepDelay)) <= 2) %>%  #for each carrier get the 2 highest delayed departures
  arrange(UniqueCarrier, desc(DepDelay))     #sorty by carrier and descending order of delays


#simpler version using `top_n` 
# http://rpackages.ianhowson.com/cran/dplyr/man/top_n.html
#wrapper that uses `filter` and `min_rank` to select the top(largest) n entries in each group, ordered by wt
flights %>%
  group_by(UniqueCarrier) %>%
  select(Month, DayofMonth, DepDelay) %>%
  top_n(2, DepDelay) %>%  
  arrange(UniqueCarrier, desc(DepDelay))


#for each month, calculate the # of flights and the change from the prev month
flights %>%
  group_by(Year, Month) %>%
  summarise(flight_count = n()) %>%
  arrange(Year, Month) %>%
  mutate(change = flight_count - lag(flight_count, n=1))

#rewrite more simply with the `tally` function
flights %>%
  group_by(Year, Month) %>%
  tally() %>%
  arrange(Year, Month) %>%
  mutate(change = n - lag(n))

```

#Other Useful Convenience Functions
* `sample_n`
* `sample_frac`

```{r other useful functions}
#randomly sample a fixed number of rows without replacement
flights %>% sample_n(size = 3, replace = FALSE)

#randomly sample a fraction(25%) of rows with replacement
flights %>% sample_frac(size = 0.25, replace=TRUE)

```


#Joining(merging) tables
* left_join
* right_join
* inner_join
* full_join
* semi_join
* anti_join
* Mutating joins
* Filtering joins
* Set operations

```{r dplyr-joins}

# create two simple data frames
a <- data_frame(color = c("green", "yellow", "red"), 
                num = 1:3)
a

b <- data_frame(color = c("green", "yellow", "pink"),
                size = c("S", "M", "L"))

b

# only include observations found in both 'a' and 'b' (automatically joins on the variables that appear in both a and b)
inner_join(a, b)

# include observations in either 'a' or 'b'
full_join(a, b)

# include all observations found in 'b'
right_join(a, b)

# right_join(a, b) is identical to left_join(b, a) except for column ordering
left_join(b, a)

# filter 'a' to only show observations that match 'b'
semi_join(a, b)

# filter 'a' to only how observations that don't match 'b'
anti_join(a, b)

# sometimes matching variables don't have identical names
b <- b %>% rename( col = color)
b

# specify that the join should occur by matching "color" in 'a' with 'col' in 'b'
inner_join(a, b, by = c("color" = "col"))

# or
a %>% inner_join(b, by = c("color" = "col"))

# Mutating Joins - Combine variables from multiple tables
flights2 <- nyflights %>% select(year:day, hour, origin, dest, tailnum, carrier)
flights2 %>% head(2)
airlines %>% head(2)

# left join by carrier
flights2 %>% left_join(airlines) %>% head(5)

# controlling how the tables are matched
weather %>% head(3)

# by default dplyr will use all common variables (a natural join)
flights2 %>% left_join(weather) %>% head(3)

# by a specified char vector (list of cols)
# note that the year columns are disambiguated with a suffix
flights2 %>%  left_join(planes, by =  "tailnum") %>% head(3)

# a named char vector
airports %>% head(2)
flights2 %>% left_join(airports, by = c("dest" = "faa")) %>% head(3)
flights2 %>% left_join(airports, by = c("origin" = "faa")) %>% head(3)


# set operations
# `intersect(x, y)`: return only observations in both x and y
df1 <- data_frame(x = 1:2, y = c(1L, 1L))
df1

df2 <- data_frame(x = 1:2, y = 1:2)
df2

# Intersection
intersect(df1, df2)

# Union
union(df1, df2)

# set difference - records in df1 but not in df2
setdiff(df1, df2)

setdiff(df2, df1)

```


#Connecting to Databases
* dplyr can connect to a database as if the data was loaded into a data frame
* Use the same syntax for local data frames and databases
* Only generates SELECT statements
* Currently supports SQLite, PostgreSQL/Redshift, MySQL/MariaDB, BigQuery, MonetDB
* Example below is based upon an SQLite database containing the hflights data
* ref: http://www3.nd.edu/~steve/computing_with_data/24_dplyr/dplyr.html

```{r dplyr-database connections}

#setup a new sqlite connection. Its just a connection with no data
my_db <- src_sqlite(path = "my_db.sqlite3", create = TRUE)


#This copies the hflights df and creates indices on the day, carrier and tailnumber to aid searching on these variables. 
#hflights_sqlite is a table object that behaves like a data.frame table but is connected to the SQLite database created on the disk.
hflights_sqlite <- copy_to(my_db, hflights, temporary = FALSE, 
                           indexes = list(c("Year", "Month", "DayofMonth"), "UniqueCarrier", "TailNum"))

class(hflights_sqlite)

hflights_sqlite

#query to get departure delays in descending order
hflights_sqlite %>%
  select(UniqueCarrier, DepDelay) %>%
  arrange(desc(DepDelay))

#get the SQL from dplyr
hflights_sqlite %>%
  select(UniqueCarrier, DepDelay) %>%
  arrange(desc(DepDelay)) %>%
  explain()

# send SQL commands to the database
tbl(my_db, sql("SELECT * FROM hflights LIMIT 5"))

```

#Creating data frames: data_frame()

`data_frame()` is a better way than data.frame() for creating data frames. Benefits of data_frame():

* You can use previously defined columns to compute new columns.
* It never coerces column types.
* It never munges column names.
* It never adds row names.
* It only recycles length 1 input.
* It returns a local data frame (a `tbl_df`).

```{r dplyr-data_frame}

# `data_frame()` example
data_frame(a = 1:6, b = a*2, c = "string", "d+e" = 1) %>% glimpse()

# data.frame() example
data.frame(a = 1:6, c = 'string', 'd+e' = 1) %>% glimpse()
# notice that variable `c` has been coerced to a `factor`

```


